---
title: "Alcohol Use, Psychological Distress, and Demographic Patterns (NSDUH2024)"
author: "Yijing Sun"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
date: "2026-04-25"
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(DT)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r data_import, include=FALSE}
 
load("data/NSDUH_2024.RData")
ls()
```

```{r data_cleaning}

nsduh_clean <- NSDUH_2024 %>%
  select(ALCDAYS, KSSLR6MONED, INCOME, IRSEX) %>%
  mutate(
    INCOME = as.numeric(as.character(INCOME)),
    IRSEX = as.numeric(as.character(IRSEX)),
    ALCDAYS = as.numeric(as.character(ALCDAYS)),
    KSSLR6MONED = as.numeric(as.character(KSSLR6MONED))
  ) %>%
  mutate(
    ALCDAYS = ifelse(ALCDAYS > 30, NA, ALCDAYS),
    KSSLR6MONED = ifelse(KSSLR6MONED > 24, NA, KSSLR6MONED),
    SEX = case_when(
      IRSEX == 1 ~ "Male",
      IRSEX == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    SEX = as.factor(SEX),
    INCOME = case_when(
      INCOME == 1 ~ "Less than\n$20,000",
      INCOME == 2 ~ "$20,000 -\n$49,999",
      INCOME == 3 ~ "$50,000 -\n$74,999",
      INCOME == 4 ~ "$75,000\nor More",
      TRUE ~ NA_character_
    ),
    INCOME = factor(
      INCOME,
      levels = c(
        "Less than\n$20,000",
        "$20,000 -\n$49,999",
        "$50,000 -\n$74,999",
        "$75,000\nor More"
      )
    )
  ) %>%
  drop_na(ALCDAYS, KSSLR6MONED, SEX, INCOME) %>%
  group_by(SEX) %>%
  mutate(
    mean_distress = mean(KSSLR6MONED, na.rm = TRUE),
    mean_alc_days = mean(ALCDAYS, na.rm = TRUE)
  ) %>%
  ungroup()

```

Main Analysis
=====================================

Row {data-height=40}
-----------------------------------------------------------------------

### Main Research Question

**How is psychological distress associated with alcohol use, and do these patterns differ by sex and income level?**

Row {data-height=500}
-------------------------------------

### Alcohol Use and Mental Health Distress

```{r plot_1}
p1 <- ggplot(
  nsduh_clean,
  aes(
    x = ALCDAYS,
    y = KSSLR6MONED,
    color = SEX,
    text = paste0(
      "<b>Sex:</b> ", SEX,
      "<br><b>Alcohol Use Days:</b> ", round(ALCDAYS, 1),
      "<br><b>Psychological Distress Score:</b> ", round(KSSLR6MONED, 1),
      "<br><b>Mean Alcohol Use Days by Sex:</b> ", round(mean_alc_days, 2),
      "<br><b>Mean Distress Score by Sex:</b> ", round(mean_distress, 2)
    )
  )
) +
  geom_point(alpha = 0.35) +
  labs(
    title = "Association Between Psychological Distress and Alcohol Use",
    x = "Days of Alcohol Use in the Past 30 Days",
    y = "Psychological Distress Score (0–24)",
    color = "Sex"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p1, tooltip = "text")
```



### Alcohol Use by Income Level

```{r plot2}

income_data <- nsduh_clean %>%
  filter(!is.na(ALCDAYS), !is.na(INCOME)) %>%
  droplevels()

anova_res <- aov(ALCDAYS ~ INCOME, data = income_data)

p_val <- summary(anova_res)[[1]][["Pr(>F)"]][1]

income_summary <- income_data %>%
  group_by(INCOME) %>%
  summarise(
    mean_alc = mean(ALCDAYS, na.rm = TRUE),
    sd_alc = sd(ALCDAYS, na.rm = TRUE),
    n = n(),
    se = sd_alc / sqrt(n),
    ci_lower = mean_alc - (1.96 * se),
    ci_upper = mean_alc + (1.96 * se),
    .groups = "drop"
  )

p2 <- ggplot(
  income_summary,
  aes(
    x = INCOME,
    y = mean_alc,
    fill = INCOME,
    text = paste0(
      "<b>Income Category:</b> ", INCOME,
      "<br><b>Mean Alcohol Days:</b> ", round(mean_alc, 2),
      "<br><b>95% CI:</b> [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]",
      "<br><b>Number of Respondents:</b> ", n,
      "<br>---",
      "<br><b>ANOVA p-value:</b> ", format.pval(p_val, digits = 3)
    )
  )
) +
  geom_col(alpha = 0.8) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    width = 0.2,
    color = "black"
  ) +
  labs(
    title = "Average Alcohol Use Days by Household Income",
    x = "Household Income Category",
    y = "Mean Days of Alcohol Use (Past 30 Days)",
    fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
    legend.position = "none",
    plot.margin = margin(t = 10, r = 20, b = 20, l = 20)
  )

ggplotly(p2, tooltip = "text")

```

Row {data-height=70}
-------------------------------------

### Real-world impact and importance of dashboard

This dashboard shows that alcohol use is associated with psychological distress and that differences across income levels are statistically significant (ANOVA p < 0.05). These findings can inform targeted public health strategies, though further modeling is needed to understand the direction and drivers of these differences.

Row {data-height=70}
-------------------------------------

### Source codes

View the full source code on GitHub:  
https://github.com/charsun0119/DATA555_final


Dataset Description
=====================================

### Dataset Description

The data used in this dashboard come from the publicly available Substance Abuse and Mental Health Services Administration National Survey on Drug Use and Health (NSDUH) 2024 dataset. The NSDUH is a nationally representative, cross-sectional survey of the U.S. civilian, noninstitutionalized population aged 12 and older, with approximately 70,000 respondents in 2024. Data are collected through a multistage probability sample using in-person and web-based interviews conducted at respondents’ residences. The study population includes individuals living in households and noninstitutional group quarters across the United States. Data were collected throughout the 2024 calendar year.
link: https://www.samhsa.gov/data/data-we-collect/nsduh-national-survey-drug-use-and-health/datafiles


Limitation
=====================================

The analysis is based on cross-sectional survey data, and so causal relationships cannot be established only based on this dataset. Additionally, the results are self-reported responses, which may be subject to bias. Moreover, only the 2024 results were involved in this dashboard, limiting the ability to assess trends over time or generalize findings beyond this specific dataset.

